---
title: "Null-model algorithms"
author: "Neander Heming, Flávio Mota and Gabriela Alves-Ferreira"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Null-model algorithms}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
link-citations: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Contents

-   [Algorithms](#algorithms)
-   [References](#references)

## Algorithms {#algorithms}

Gotelli (2000) summarizes nine null model algorithms for species co-occurrence analysis based on how sums of species (originally rows) and sites (originally columns) are treated (i.e. fixed, equiprobable, or proportional sums; see Table \@ref(tab:table1), [@gotelli2000]). The data for null model analyses usually consists of a binary presence-absence matrix, in which rows represent species or taxa, columns represent sites or samples, and the entries represent the presence (1) or absence (0) of a particular species in a particular site [@ulrich2012].

```{r table1-data, echo=FALSE}
type <- c("Equiprobable", "Proportional", "Fixed")
typeabv <- c("E", "P", "F") #c("Equipr.", "Prop.", "Fixed")
typeinit <- c("E", "P", "F")
# Species - Sites
d <- data.frame("Rows" = c("Rows (Layer)", rep("", 2)),
                type=type,
                paste0(
                  paste("Spp occur.:", typeabv, 
                        "<br> Site rich.:", typeabv[1]),
                  "<br>", "SIM", c(1,7,2), ": ", typeinit, typeinit[1], 
                  "") ,
                paste0(
                  paste("Spp occur.:", typeabv, 
                        "<br> Site rich.:", typeabv[2]), 
                  "<br>", "SIM", c(6,8,4), ": ", typeinit, typeinit[2], 
                  "") ,
                paste0(
                  paste("Spp occur.:", typeabv, 
                        "<br> Site rich.:", typeabv[3]), 
                  "<br>", "SIM", c(3,5,9), ": ", typeinit, typeinit[3], 
                  "") )
colnames(d) <- c(" ", " ", paste0("", type))
```

```{r table1, echo=FALSE}
require(kableExtra)

kable(d, dbooktabs = TRUE, align=c("r", "c", "c", "c", "c"), escape = F, 
      caption = "Nine null model algorithms for species co-occurrence analysis listed in Gotelli (2000). Algorithms marked in green are already implemented in SESraster") %>%
  collapse_rows() %>%
  kable_styling(bootstrap_options = c("bordered")) %>%
  add_header_above(c(" " = 2, "Cols (Cell)" = 3), background = "#F2F2F2") %>%
  row_spec(0, background = "#F2F2F2") %>% # , extra_css = c("border-bottom-style: none", "", "", "")
  column_spec(1:2, bold = TRUE, background="#F2F2F2", border_right = TRUE) %>%
  column_spec(3:4, border_right = TRUE) %>%
  column_spec(3, background=c("#B4EEB4")) %>%
  column_spec(4, background=c("#FFFAFA")) %>%
  column_spec(5, background=c("#B4EEB4"))

```

## References {#references}
