---
title: "SES calculation"
author: "Neander M. Heming, Flávio Mota, and Gabriela Alves-Ferreira"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SES calculation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
# csl: restoration-ecology.csl
# biblio-style: apelike
link-citations: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(SESraster)
```

## Contents

*   [Standardized effect sizes (SES)](#ses)
*   [Calculating SES](#analysis)
    + [Random species generation](#randsp)
    + [SES with spatial randomization](#ses-spat)
    + [SES from species trait randomization](#ses-trait)
*   [References](#references)

## Standardized effect sizes (SES) {#ses}
Standardized effect sizes (SES) are....

## Calculating SES {#analysis}
### Random species generation {#randsp}

Now, let's see some of the package features. First, we will create some random species distributions using the package `terra`.

```{r, rand-spp, fig.height=4, fig.width=4}
library(SESraster)
library(terra)
# creating random species distributions
f <- system.file("ex/elev.tif", package="terra")
r <- rast(f)
set.seed(510)
r <- rast(lapply(1:18,
                function(i, r, mn, mx){
                  app(r, function(x, t){
                    sapply(x, function(x, t){
                       x<max(t) & x>min(t)
                    }, t=t)
                  }, t=sample(seq(mn, mx), 2))
                }, r=r, mn=minmax(r)[1]+10, mx=minmax(r)[2]-10))

names(r) <- paste("sp", 1:nlyr(r))
plot(r)
```

With the distributions in hand, we can perform the spatial randomizations.


### SES with spatial randomization {#ses-spat}

First we need a function that computes the desired metric. The function must work
with spatial data. Just to exemplify, we are creating a function to compute the mean of
presences and absences (1/0) within each cell. You probably wants to use a more 
ecologically meaningful function, but here is just an example of use.

```{r, ses-fun}
appmean <- function(x, ...){
                      terra::app(x, "mean", ...)
}
```

Now, to compute SES, we will send our function `appmean()` to `SESraster()` through
'FUN' argument to compute our metric. We also randomize the original data using
the `bootspat_naive()` algorithm and randomize by 'species' passing the argument 
'random="species"' through 'spat_alg_args'.

```{r, ses-spat-sp, fig.height=4, fig.width=4}
ses.sp <- SESraster(r, FUN=appmean, 
                    spat_alg = "bootspat_naive", spat_alg_args=list(random="species"),
                    aleats = 5)
plot(ses.sp)
```

Compute metric and SES using `bootspat_naive()` and randomize by site.

```{r, ses-spat-st, fig.height=4, fig.width=4}
ses.st <- SESraster(r, FUN=appmean,
                    spat_alg = "bootspat_naive", spat_alg_args=list(random="site"),
                    aleats = 5)
plot(ses.st)
```


#### Passing arguments to 'FUN' {#ses-fun-arg}

Using 'FUN_args' to send arguments to the function that calculates the desired metric ('FUN')

```{r, ses-spat-na, fig.height=4, fig.width=4}

# r[7][1] <- NA
# plot(r)
sesNA <- SESraster(r, FUN=appmean, FUN_args = list(na.rm = FALSE),
                   spat_alg = "bootspat_naive", spat_alg_args=list(random = "species"),
                   aleats = 5)
plot(sesNA)
#'
ses.woNA <- SESraster(r, FUN=appmean, FUN_args = list(na.rm = TRUE), 
                      spat_alg = "bootspat_naive", spat_alg_args=list(random = "species"),
                      aleats = 5)
plot(ses.woNA)
```

### SES from species trait randomization {#ses-trait}

```{r, ses-trait, fig.height=4, fig.width=4}
## example with 'Fa_alg'
appsv <- function(x, lyrv, na.rm = FALSE, ...){
                      sumw <- function(x, lyrv, na.rm, ...){
                            ifelse(all(is.na(x)), NA,
                                    sum(x*lyrv, na.rm=na.rm, ...))
                      }
                      terra::app(x, sumw, lyrv=lyrv, na.rm, ...)
                    }
ses <- SESraster(r, FUN=appsv,
                 FUN_args = list(lyrv = seq_len(nlyr(r)), na.rm = TRUE),
                    Fa_sample = "lyrv",
                    Fa_alg = "sample", Fa_alg_args = list(replace=FALSE),
                    aleats = 5)
plot(ses)

ses <- SESraster(r, FUN=appsv,
                 FUN_args = list(lyrv = seq_len(nlyr(r)), na.rm = TRUE),
                    Fa_sample = "lyrv",
                    Fa_alg = "sample", Fa_alg_args = list(replace=TRUE),
                    aleats = 5)
plot(ses)
```

## References {#references}
